---
title: "Technical Integration - Using textAnnotatoR with Other R Tools"
author: "Chao Liu"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Technical Integration - Using textAnnotatoR with Other R Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = 'center',
  fig.path = "figures/"
)
library(textAnnotatoR)
```

# Introduction

While textAnnotatoR provides a comprehensive GUI for annotation, its true power comes from integration with the broader R ecosystem. This vignette demonstrates how to integrate textAnnotatoR's outputs with other popular R packages for text analysis, visualization, and reporting.

# Understanding Data Structures

Before integrating with other tools, let's explore the structure of textAnnotatoR's data:

## Project Files

textAnnotatoR project files (.rds) contain the following components:

```{r echo = T, results = 'hide'}
# First, ensure the project directory exists and is initialized
# We need to handle this explicitly when working outside the GUI
initialize_project_directory <- function() {
  # Get user documents directory as a safer default
  user_docs <- file.path(path.expand("~"), "Documents")
  
  # Create textAnnotatoR directory if it doesn't exist
  text_annotator_dir <- file.path(user_docs, "textAnnotatoR")
  if (!dir.exists(text_annotator_dir)) {
    dir.create(text_annotator_dir, recursive = TRUE)
  }
  
  # Create projects subdirectory
  projects_dir <- file.path(text_annotator_dir, "projects")
  if (!dir.exists(projects_dir)) {
    dir.create(projects_dir, recursive = TRUE)
  }
  
  return(projects_dir)
}

# Initialize the directory
projects_dir <- initialize_project_directory()

# Now load a project with an absolute path
project_path <- file.path(projects_dir, "my_project.rds")

# If you're loading an existing project file from another location,
# you may need to copy it to the projects directory first
if (file.exists("C:/Users/chaoliu/Downloads/my_project.rds")) {
  file.copy("C:/Users/chaoliu/Downloads/my_project.rds", project_path)
}

# Load the project
if (file.exists(project_path)) {
  project <- readRDS(project_path)
} else {
  # If project doesn't exist, create a sample one for demonstration
  project <- list(
    text = "Sample interview text content...",
    annotations = data.frame(
      start = c(1, 50),
      end = c(10, 60),
      text = c("Sample", "interview"),
      code = c("sample_code", "another_code"),
      memo = c("Sample memo", "Another memo"),
      stringsAsFactors = FALSE
    ),
    codes = c("sample_code", "another_code"),
    code_tree = data.tree::Node$new("Root"),
    code_colors = c("sample_code" = "#FF0000", "another_code" = "#00FF00"),
    memos = list(),
    code_descriptions = list()
  )
  
  # Save the sample project
  saveRDS(project, project_path)
  cat("Created sample project at:", project_path, "\n")
}

# Structure of a project
str(project, max.level = 1)
#> List of 9
#>  $ text             : chr "Interviewer: Can you describe your experience transitioning to remote work?\n\nParticipant: It was definitely c"| __truncated__
#>  $ annotations      :'data.frame':	2 obs. of  5 variables:
#>  $ codes            : chr [1:2] "Home Setup" "Boundary difficulty"
#>  $ code_tree        :Classes 'Node', 'R6' <Node>
#>  ...
```

## Annotation Data Frame

The core annotation data is stored in a standard R data frame:

```{r eval=TRUE}
# View annotation structure
head(project$annotations)
#>   start   end                              text        code                   memo
#> 1   125   174 setting up a proper workspace at home Home Setup Initial challenge in...
#> 2   179   248 establishing boundaries between work... Work-Life Recurring theme in...
```

This structure makes it readily compatible with standard R data manipulation tools.

# Integrating with Text Analysis Packages

## Using textAnnotatoR with quanteda

The [quanteda](https://quanteda.io/) package is a powerful framework for text analysis. Here's how to combine textAnnotatoR annotations with quanteda:

```{r message=FALSE, warning=FALSE}
library(quanteda)
library(quanteda.textstats)
library(dplyr)
```

```{r eval=TRUE}
# Extract annotations by code
home_setup_texts <- project$annotations %>%
  filter(code == "Home Setup") %>%
  pull(text)

# Create a corpus from these text segments
home_corpus <- corpus(home_setup_texts)

# Analyze with quanteda
home_dfm <- dfm(tokens(home_corpus))
textstat_frequency(home_dfm, n = 10)  # Top 10 words
```

## Integration with tidytext

The [tidytext](https://www.tidytextmining.com/) package works beautifully with textAnnotatoR outputs:

```{r message=FALSE, warning=FALSE}
library(tidytext)
library(ggplot2)
```


```{r eval = TRUE}
# Convert annotations to tidy format
tidy_annotations <- project$annotations %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words)

# Word frequency by code
word_frequencies <- tidy_annotations %>%
  count(code, word, sort = TRUE) %>%
  group_by(code) %>%
  top_n(10, n)

# Visualize
ggplot(word_frequencies, aes(reorder(word, n), n, fill = code)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~code, scales = "free") +
  coord_flip() +
  labs(title = "Top 10 Words by Code",
       x = NULL,
       y = "Count")
```

# Enhanced Visualization

## Network Visualization with igraph

textAnnotatoR's co-occurrence data can be visualized as networks:

```{r message=FALSE, warning=FALSE}
library(igraph)
library(ggraph)
```


```{r eval = TRUE}
# Create a matrix of code co-occurrences
# (This code assumes you've run analyze_co_occurrences or similar)
co_matrix <- matrix(c(
  0, 3, 1, 0,
  3, 0, 2, 4,
  1, 2, 0, 2,
  0, 4, 2, 0
), nrow = 4, byrow = TRUE)

rownames(co_matrix) <- colnames(co_matrix) <- 
  c("Home Setup", "Work-Life", "Tools", "Benefits")

# Create a graph
g <- graph_from_adjacency_matrix(
  co_matrix,
  mode = "undirected",
  weighted = TRUE,
  diag = FALSE
)

# Plot with ggraph
ggraph(g, layout = "fr") +
  geom_edge_link(aes(width = weight), alpha = 0.7) +
  geom_node_point(size = 7, color = "lightblue") +
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_void() +
  labs(title = "Code Co-occurrence Network")
```

## Interactive Visualizations with plotly

Make your textAnnotatoR analyses interactive with plotly:

```{r message=FALSE, warning=FALSE}
library(plotly)

# Create a base ggplot with code frequencies
code_freq <- data.frame(
  code = project$codes,
  frequency = sapply(project$codes, function(c) sum(project$annotations$code == c))
)

p <- ggplot(code_freq, aes(reorder(code, frequency), frequency, fill = code)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(title = "Code Frequencies", x = "Code", y = "Frequency")

# Convert to interactive plot
ggplotly(p) %>%
  layout(hoverlabel = list(bgcolor = "white"))
```

# Batch Processing and Automation

## Processing Multiple Documents

For projects with multiple documents, you can automate the workflow:

```{r eval=FALSE}
# Function to process a single document
process_document <- function(file_path, coding_scheme) {
  # Import text
  text <- readLines(file_path, warn = FALSE)
  text <- paste(text, collapse = "\n")
  
  # Create a new project structure
  project <- list(
    text = text,
    annotations = data.frame(
      start = integer(),
      end = integer(),
      text = character(),
      code = character(),
      memo = character(),
      stringsAsFactors = FALSE
    ),
    codes = coding_scheme$codes,
    code_tree = coding_scheme$code_tree,
    code_colors = coding_scheme$code_colors,
    memos = list(),
    code_descriptions = coding_scheme$code_descriptions
  )
  
  # Save as a textAnnotatoR project
  project_name <- paste0(basename(tools::file_path_sans_ext(file_path)), ".rds")
  save_project_state(project, project_name)
  
  cat("Processed:", file_path, "\n")
  return(project_name)
}

# Process a folder of documents
files <- list.files("interview_transcripts", pattern = "*.txt", full.names = TRUE)
processed_projects <- sapply(files, process_document, coding_scheme = master_scheme)
```

## Export to Other Formats

Convert textAnnotatoR annotations to formats compatible with other QDA software:

```{r eval=FALSE}
# Export to REFI-QDA format (compatible with many QDA tools)
export_to_refi_qda <- function(project, output_file) {
  # This would need to be implemented based on REFI-QDA specifications
  # See: https://www.qdasoftware.org/
  
  # Basic structure
  xml_content <- paste0(
    '<?xml version="1.0" encoding="UTF-8"?>\n',
    '<qdpx:project xmlns:qdpx="www.qdasoftware.org">\n',
    '  <qdpx:sources>\n',
    '    <qdpx:textSource>\n',
    '      <qdpx:plainTextContent>', 
    project$text,
    '      </qdpx:plainTextContent>\n',
    '    </qdpx:textSource>\n',
    '  </qdpx:sources>\n',
    '  <qdpx:codes>\n'
  )
  
  # Add codes
  for (code in project$codes) {
    xml_content <- paste0(
      xml_content,
      '    <qdpx:code name="', code, '" />\n'
    )
  }
  
  # Complete XML structure
  xml_content <- paste0(
    xml_content,
    '  </qdpx:codes>\n',
    '</qdpx:project>'
  )
  
  # Write to file
  writeLines(xml_content, output_file)
  
  return(output_file)
}
```

# Integration with Reporting Tools

## Creating Reports with R Markdown

Generate comprehensive reports from your textAnnotatoR analyses:

```{r message=FALSE, warning=FALSE}
library(rmarkdown)

# Function to generate a report
generate_analysis_report <- function(project, output_file = "analysis_report.html") {
  # Create a temporary Rmd file
  temp_rmd <- tempfile(fileext = ".Rmd")
  
  # Write R Markdown content
  writeLines(
    c("---",
      "title: 'Qualitative Analysis Report'",
      "output: html_document",
      "---",
      "",
      "```{r setup, include=FALSE}",
      "knitr::opts_chunk$set(echo = FALSE)",
      "library(ggplot2)",
      "library(DT)",
      "```",
      "",
      "# Analysis Overview",
      "",
      "This report presents findings from qualitative analysis using textAnnotatoR.",
      "",
      "## Document Statistics",
      "",
      "```{r}",
      "stats <- data.frame(",
      "  Metric = c('Total Words', 'Total Annotations', 'Unique Codes'),",
      paste0("  Value = c(length(strsplit(project$text, '\\\\W+')[[1]]), nrow(project$annotations), length(project$codes))"),
      ")",
      "knitr::kable(stats)",
      "```",
      "",
      "## Code Frequency",
      "",
      "```{r fig.height=6, fig.width=8}",
      "code_counts <- table(project$annotations$code)",
      "code_df <- data.frame(",
      "  Code = names(code_counts),",
      "  Frequency = as.numeric(code_counts)",
      ")",
      "ggplot(code_df, aes(reorder(Code, Frequency), Frequency, fill = Code)) +",
      "  geom_col() +",
      "  coord_flip() +",
      "  theme_minimal() +",
      "  labs(title = 'Code Frequencies', x = 'Code', y = 'Frequency') +",
      "  theme(legend.position = 'none')",
      "```",
      "",
      "## Annotations Table",
      "",
      "```{r}",
      "DT::datatable(project$annotations)",
      "```"
    ),
    temp_rmd
  )
  
  # Render the report
  rmarkdown::render(temp_rmd, output_file = output_file)
  
  return(output_file)
}
```

## Integration with Shiny

Create interactive dashboards from your textAnnotatoR projects:

```{r message=FALSE, warning=FALSE}
library(shiny)

# A simple Shiny app to explore annotations
launch_annotation_explorer <- function(project) {
  ui <- fluidPage(
    titlePanel("textAnnotatoR Explorer"),
    
    sidebarLayout(
      sidebarPanel(
        selectInput("code", "Select Code:",
                    choices = c("All", project$codes)),
        hr(),
        checkboxGroupInput("display_cols", "Display Columns:",
                           choices = colnames(project$annotations),
                           selected = c("text", "code", "memo"))
      ),
      
      mainPanel(
        tabsetPanel(
          tabPanel("Annotations", DTOutput("annotations_table")),
          tabPanel("Code Frequencies", plotOutput("code_freq_plot")),
          tabPanel("Text", verbatimTextOutput("text_display"))
        )
      )
    )
  )
  
  server <- function(input, output) {
    # Filter annotations based on selected code
    filtered_annotations <- reactive({
      if (input$code == "All") {
        project$annotations
      } else {
        project$annotations[project$annotations$code == input$code, ]
      }
    })
    
    # Display annotations table
    output$annotations_table <- renderDT({
      filtered_annotations()[, input$display_cols, drop = FALSE]
    })
    
    # Display code frequency plot
    output$code_freq_plot <- renderPlot({
      code_counts <- table(project$annotations$code)
      code_df <- data.frame(
        Code = names(code_counts),
        Frequency = as.numeric(code_counts)
      )
      
      ggplot(code_df, aes(reorder(Code, Frequency), Frequency, fill = Code)) +
        geom_col() +
        coord_flip() +
        theme_minimal() +
        labs(title = "Code Frequencies", x = "Code", y = "Frequency") +
        theme(legend.position = "none")
    })
    
    # Display text with highlighted segments
    output$text_display <- renderText({
      if (input$code == "All") {
        project$text
      } else {
        # Get annotations for the selected code
        code_annotations <- project$annotations[project$annotations$code == input$code, ]
        
        # Sort by position
        code_annotations <- code_annotations[order(code_annotations$start), ]
        
        # Extract and display text segments
        segments <- sapply(1:nrow(code_annotations), function(i) {
          start <- code_annotations$start[i]
          end <- code_annotations$end[i]
          paste0("* ", substr(project$text, start, end))
        })
        
        paste(segments, collapse = "\n\n")
      }
    })
  }
  
  shinyApp(ui, server)
}
```

# Performance Tips

## Working with Large Documents

For very large documents, consider these optimization strategies:

```{r eval=FALSE}
# 1. Split large documents
split_large_document <- function(file_path, chunk_size = 5000) {
  text <- readLines(file_path, warn = FALSE)
  text <- paste(text, collapse = "\n")
  
  # Split into chunks
  chunks <- list()
  for (i in seq(1, nchar(text), by = chunk_size)) {
    end_pos <- min(i + chunk_size - 1, nchar(text))
    chunks[[length(chunks) + 1]] <- substr(text, i, end_pos)
  }
  
  # Save chunks
  chunk_files <- character(length(chunks))
  for (i in seq_along(chunks)) {
    filename <- paste0(tools::file_path_sans_ext(basename(file_path)), 
                      "_chunk", i, ".txt")
    writeLines(chunks[[i]], filename)
    chunk_files[i] <- filename
  }
  
  return(chunk_files)
}

# 2. Optimize annotations
optimize_annotations <- function(project) {
  # Remove duplicate annotations
  project$annotations <- unique(project$annotations)
  
  # Sort by position
  project$annotations <- project$annotations[order(project$annotations$start), ]
  
  return(project)
}
```

## Batch Processing Tips

When working with multiple files or team members:

```{r eval=FALSE}
# Function to merge projects from multiple team members
merge_team_projects <- function(project_files, output_file = "merged_project.rds") {
  # Load all projects
  projects <- lapply(project_files, load_project_state)
  
  # Ensure all projects refer to the same text
  text <- projects[[1]]$text
  if (!all(sapply(projects, function(p) identical(p$text, text)))) {
    stop("All projects must reference the same text document")
  }
  
  # Collect all annotations
  all_annotations <- do.call(rbind, lapply(projects, function(p) {
    # Add a coder column to track source
    p$annotations$coder <- deparse(substitute(p))
    p$annotations
  }))
  
  # Create merged project
  merged_project <- projects[[1]]
  merged_project$annotations <- all_annotations
  
  # Save merged project
  save_project_state(merged_project, output_file)
  
  return(merged_project)
}
```

# Conclusion

This vignette demonstrates how to leverage textAnnotatoR's data structures to integrate with the broader R ecosystem. By combining the intuitive annotation interface with powerful R packages for analysis and visualization, you have more control over your qualitative analysis workflow.

---

*Note: Some functions in this vignette are examples that may need modification based on your specific needs and the current version of textAnnotatoR.*
